A \texttt{shiny} New Opportunity for Interaction with Big Data in Undergraduate Education

### 2015-07-07 Body/ShinyDBSampler/ShinyDBSampler.Rnw
Finally got the page numbers to fall in right spot in top center (thanks to Susan!). Also, LOTS of table and figure caption formatting to meet ISU thesis specifications.  Chapter 1, 2 and 3 completed.

\begin{figure}[hbtp]
\centering
\includegraphics[keepaspectratio=TRUE,width=.95\textwidth]{ShinyDBSampler/SampleAndSummarizeWithPanelLabels.png}
\caption{Shiny Database Sampler Layout for \textit{Sample and Summarize} Tab} 
\label{fig:samplesummarizetab}
\end{figure}

@@

Once the plotting options are selected in the sidebar panel, the user may click the \textit{Make my Plot!} button to generate and display the plot in the main panel. Since the visualizations are intended only for preliminary exploration, the plots have a default construction and labeling that are not able to be customized.

\begin{figure}[htbp]
\begin{center}
\includegraphics[keepaspectratio=TRUE,width=.99\textwidth]{ShinyDBSampler/Visualize.png}
\end{center}
\caption{Shiny Database Sampler Layout for \textit{Visualize} Tab} 
\label{fig:visualizetab}
\end{figure}

\begin{table}[hbtp]
\centering
\begin{tabular}{lll}
Response Variable & Explanatory Variable & Default Plot \\
\hline
Numerical & None & Histogram \\
Categorical & None & Barchart \\
Numerical & Numerical & Scatterplot \\
Categorical & Categorical & Stacked Barchart \\
Numerical & Categorical & Side-by-side Boxplots \\
\end{tabular}
\caption{Plot types supported in \textit{Visualize} tab.} 
\label{tab:defaultplottypes}
\end{table}

@@

After completing the lab assignment, students were asked to respond to a survey consisting of twelve statements (referred to as \textit{items} in the following, see Table~\ref{tab:surveyquestions} for an overview). For each statement, students were asked for feedback on their level of agreement on a Likert scale from strongly disagree to strongly agree. 
The twelve items were designed to assess student opinion within three topics of four items each: ease of use, connection to sampling concepts, and engagement with the census data.  We will refer to these as the \textit{Ease}, \textit{Concept} and \textit{Engagement} item sets.  For each group of four items, two were worded positively and two were worded negatively. Introducing negation with half of the items was done to reduce the response bias associated with \textit{acquiescence} -- the tendency to respond positively irrespective of the item content due \citep{Furnham1986}. Responses were scored as -2 (strongly disagree), -1 (disagree), 0 (neutral), 1 (agree), 2 (strongly agree).  Responses for negatively worded items were reverse-scored for the purposes of analysis.

\begin{table}[hbtp]
\centering
\small
% \scalebox{0.9}{
\begin{tabular}{ccp{3.4in}crr}
Topic Set & ID & Item & Polarity & Mean & SD\\[3pt]
  \hline
Ease & 1 & \it I found the web tool easy to use & $+$ & \Sexpr{sprintf("%.2f", sumstats[1,2])} & \Sexpr{sprintf("%.2f", sumstats[1,3])}\\  
 & 2 & \it The layout of the web tool was intuitive & $+$  & \Sexpr{sprintf("%.2f", sumstats[4,2])} & \Sexpr{sprintf("%.2f", sumstats[4,3])}\\  
 & 3 & \it Using the web tool was difficult & $-$  & \Sexpr{sprintf("%.2f", sumstats[7,2])}& \Sexpr{sprintf("%.2f", sumstats[7,3])}\\  
 & 4 & \it Learning to use the web tool was hard & $-$  & \Sexpr{sprintf("%.2f", sumstats[10,2])} & \Sexpr{sprintf("%.2f", sumstats[10,3])}\\  [3pt]
  \hline
Concept & 1 & \it The web tool helped me understand sampling concepts & $+$  & \Sexpr{sprintf("%.2f", sumstats[2,2])}& \Sexpr{sprintf("%.2f", sumstats[2,3])}\\   
& 2 & \it I understand sampling ideas less after using the web tool & $-$  & \Sexpr{sprintf("%.2f", sumstats[5,2])}& \Sexpr{sprintf("%.2f", sumstats[5,3])}\\  
& 3 & \it Sampling techniques are clearer after using the web tool & $+$  & \Sexpr{sprintf("%.2f", sumstats[8,2])}& \Sexpr{sprintf("%.2f", sumstats[8,3])}\\  
& 4 & \it The web tool made me less sure how to randomly sample & $-$  & \Sexpr{sprintf("%.2f", sumstats[11,2])}& \Sexpr{sprintf("%.2f", sumstats[11,3])}\\  
   \hline
Engagement & 1 & \it I did not enjoy working with the Census data & $-$  & \Sexpr{sprintf("%.2f", sumstats[3,2])} & \Sexpr{sprintf("%.2f", sumstats[3,3])}\\  
& 2 & \it I thought the Census data was boring & $-$   & \Sexpr{sprintf("%.2f", sumstats[6,2])}& \Sexpr{sprintf("%.2f", sumstats[6,3])}\\  
& 3 & \it Knowing that the Census data was from real people made it more interesting & $+$  & \Sexpr{sprintf("%.2f", sumstats[9,2])}& \Sexpr{sprintf("%.2f", sumstats[9,3])}\\   
& 4 & \it I liked analyzing the Census data & $+$  & \Sexpr{sprintf("%.2f", sumstats[12,2])} & \Sexpr{sprintf("%.2f", sumstats[12,3])}\\   
\end{tabular}
% }
\caption{Survey questions and response summaries for all items \textit{after} Reverse-Scoring (RS)}
\label{tab:surveyquestions}
\end{table}

@@

<<fluctuationDiagrams,echo=F, message=FALSE, include=T,eval=T,fig.width=12, fig.height=6, out.width='.98\\linewidth', fig.pos='hbt',fig.align='center',tidy=F, cache=TRUE, fig.show='hold', fig.cap='Fluctuation Diagrams of All Item Pairs within Topic Sets'>>=

@@

\begin{table}[hbtp]
\centering
\begin{tabular}{ll}
\hline 
Internal Consistency & Range \\
\hline
Excellent &  $[ 0.9 , 1.0 ]$ \\
Good & $[ 0.8 , 0.9 )$ \\
Acceptable & $[ 0.7 , 0.8 ) $\\
Questionable & $[ 0.6 , 0.7 )$ \\
Poor & $[ 0.5 , 0.6 )$ \\
Unacceptable & $( -\infty, 0.5 )$ \\
\hline
\end{tabular}
\caption{Extended Scale for Cronbach's $\alpha$  (George and Mallery, 2003).} 
\label{GMAlphaScale}
\end{table}

@@

Table~\ref{cronbachstuff} displays the point estimates and 95\% central bootstrap intervals for Cronbach's $\alpha$ for each item set from the student survey.  The intervals were created using quantiles of the Cronbach's $\alpha$ values from 10,000 bootstrap resamples. 
 The results indicate modest levels of internal consistency for Ease and Engagement item sets, and a lower level for the Concept item set. This is in agreement with the findings based on the fluctuation diagrams in Figure~\ref{fig:fluctuationDiagrams}.

@@

% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Tue Nov 04 15:05:46 2014
\begin{table}[hbtp]
\centering
\begin{tabular}{lrc}
  \hline
Set & Estimate & 95\% Confidence Interval \\ 
  \hline
Ease & 0.70 & (0.613 , 0.759)\\ 
  Concept & 0.53 & (0.410 , 0.637) \\ 
  Engagement & 0.72 & (0.643 , 0.776) \\ 
   \hline
\end{tabular}
\caption{Cronbach's $\alpha$ Estimates for each item set with 95\% central bootstrap confidence interval based on 10,000 bootstrap samples} 
\label{cronbachstuff}
\end{table}

@@

<<OverallSetBars,echo=F, message=FALSE, include=T ,eval=T,fig.width=9, fig.height=2.5, out.width='.99\\linewidth', fig.pos='hbt',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap= 'Item set response distributions'>>=
# meltdatNoNA$value <- as.numeric(as.character(meltdatNoNA$value))
# qu.summary <- ddply(meltdatNoNA, .(set, question.in.set, value), summarize,
# n=length(value) )
# qu.summary$polarity <- rep(c("Pos","Pos","Neg","Neg","Pos","Neg","Pos","Neg","Neg","Neg","Pos","Pos"),each=5)
# qu.summary$glyphShift <- .05
# qu.summary$glyphShift[qu.summary$polarity=="Neg"] <- -.05
# 
# qu.means <- ddply(qu.summary, .(set, value, polarity), summarize, means=mean(n))
# 
# #write.csv(qu.means, "quMeans.csv",row.names=FALSE)
qu.means <- read.csv("quMeans.csv",header=T)

@@

\begin{table}[hbtp]
\centering
\begin{tabular}{lrrrrrrr}
\hline 
& Principal Component & 1 & 2 & 3 & 4 & 5 & 6 \\ 
\hline
Variances \hspace{.1in} & Prop. of Var & \Sexpr{sprintf("%.3f", pc6vars[1,2])} & \Sexpr{sprintf("%.3f", pc6vars[2,2])}  & \Sexpr{sprintf("%.3f", pc6vars[3,2])} & \Sexpr{sprintf("%.3f", pc6vars[4,2])} & \Sexpr{sprintf("%.3f", pc6vars[5,2])} & \Sexpr{sprintf("%.3f", pc6vars[6,2])}  \\
& Cumu. Prop. of Var & \Sexpr{sprintf("%.3f", pc6vars[1,3])} & \Sexpr{sprintf("%.3f", pc6vars[2,3])}  & \Sexpr{sprintf("%.3f", pc6vars[3,3])} & \Sexpr{sprintf("%.3f", pc6vars[4,3])} & \Sexpr{sprintf("%.3f", pc6vars[5,3])} & \Sexpr{sprintf("%.3f", pc6vars[6,3])}  \\
\hline
Loadings & Pos. Ease & \Sexpr{sprintf("%.3f", pc6loadings[1,1])} & \Sexpr{sprintf("%.3f", pc6loadings[1,2])} & \Sexpr{sprintf("%.3f", pc6loadings[1,3])} & \Sexpr{sprintf("%.3f", pc6loadings[1,4])} & \Sexpr{sprintf("%.3f", pc6loadings[1,5])} & \Sexpr{sprintf("%.3f", pc6loadings[1,6])}\\
& Neg. Ease &  \Sexpr{sprintf("%.3f", pc6loadings[2,1])} & \Sexpr{sprintf("%.3f", pc6loadings[2,2])} & \Sexpr{sprintf("%.3f", pc6loadings[2,3])} & \Sexpr{sprintf("%.3f", pc6loadings[2,4])} & \Sexpr{sprintf("%.3f", pc6loadings[2,5])} & \Sexpr{sprintf("%.3f", pc6loadings[2,6])}\\
& Pos. Concept &  \Sexpr{sprintf("%.3f", pc6loadings[3,1])} & \Sexpr{sprintf("%.3f", pc6loadings[3,2])} & \Sexpr{sprintf("%.3f", pc6loadings[3,3])} & \Sexpr{sprintf("%.3f", pc6loadings[3,4])} & \Sexpr{sprintf("%.3f", pc6loadings[3,5])} & \Sexpr{sprintf("%.3f", pc6loadings[3,6])}\\
& Neg. Concept &  \Sexpr{sprintf("%.3f", pc6loadings[4,1])} & \Sexpr{sprintf("%.3f", pc6loadings[4,2])} & \Sexpr{sprintf("%.3f", pc6loadings[4,3])} & \Sexpr{sprintf("%.3f", pc6loadings[4,4])} & \Sexpr{sprintf("%.3f", pc6loadings[4,5])} & \Sexpr{sprintf("%.3f", pc6loadings[4,6])}\\
& Pos. Engaged &  \Sexpr{sprintf("%.3f", pc6loadings[5,1])} & \Sexpr{sprintf("%.3f", pc6loadings[5,2])} & \Sexpr{sprintf("%.3f", pc6loadings[5,3])} & \Sexpr{sprintf("%.3f", pc6loadings[5,4])} & \Sexpr{sprintf("%.3f", pc6loadings[5,5])} & \Sexpr{sprintf("%.3f", pc6loadings[5,6])}\\
& Neg. Engaged &  \Sexpr{sprintf("%.3f", pc6loadings[6,1])} & \Sexpr{sprintf("%.3f", pc6loadings[6,2])} & \Sexpr{sprintf("%.3f", pc6loadings[6,3])} & \Sexpr{sprintf("%.3f", pc6loadings[6,4])} & \Sexpr{sprintf("%.3f", pc6loadings[6,5])} & \Sexpr{sprintf("%.3f", pc6loadings[6,6])}\\
\hline
\end{tabular}
\caption{Summary Statistics from Principal Component Analysis with Six Topic/Polarity Item Pairs} 
\label{pc6table}
\end{table}

<<PCA6plot,echo=F, message=FALSE, include=T,eval=T,fig.width=5, fig.height=5, out.width='.5\\linewidth', fig.pos='hbt',fig.align='center',tidy=F, cache=TRUE, fig.show='hold', fig.cap='Item Pair Loadings on Components 2 and 3 from the Principal Component Analysis with Six Topic/Polarity Item Pairs'>>=
load23dat <- data.frame(pc6loadings[,2:3], compname = row.names(pc6loadings))
require(grid)
ggplot(data=load23dat)+ 
  geom_hline(x=0, colour="darkgray") + geom_vline(y=0, colour="darkgray")+
  geom_point(aes(x=Comp.2, y=Comp.3),size=3.5) + xlim(c(-.73,.6)) + 
  theme_bw() + xlab("Component 2") + ylab("Component 3") +
  geom_segment(aes(x = 0, y = 0, xend = Comp.2, yend = Comp.3), colour="grey40") + 
  geom_text(aes(x=0.825*Comp.2, y=0.825*Comp.3, label=compname)) +coord_fixed() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="cm"))
@

@@

The next major consideration is whether the item sets are truly measuring different latent topics, and therefore can be view as non-redundant. The ability of the survey to separately measure the topics of Ease, Concepts and Engagement can be assessed through the orthogonality of the responses from different item sets. To check the orthogonality of the sets we conduct another principal component analysis; this time on the average responses for each student from the four item sets -- Ease, Positive Concept, Negative Concept and Engagement.  Items sets will be considered highly orthogonal if the principle component analysis cannot reduce the dimensionality from the four sets.

@@

\begin{table}[hbtp]
\centering
\begin{tabular}{lrrrrr}
\hline 
& Principal Component & 1 & 2 & 3 & 4 \\ 
\hline
Variances \hspace{.1in} & Prop. of Var & \Sexpr{sprintf("%.3f", pc4vars[1,2])} & \Sexpr{sprintf("%.3f", pc4vars[2,2])}  & \Sexpr{sprintf("%.3f", pc4vars[3,2])} & \Sexpr{sprintf("%.3f", pc4vars[4,2])}  \\
& Cumu. Prop. of Var & \Sexpr{sprintf("%.3f", pc4vars[1,3])} & \Sexpr{sprintf("%.3f", pc4vars[2,3])}  & \Sexpr{sprintf("%.3f", pc4vars[3,3])} & \Sexpr{sprintf("%.3f", pc4vars[4,3])}  \\
\hline
Loadings & Ease & \Sexpr{sprintf("%.3f", pc4loadings[1,1])} & \Sexpr{sprintf("%.3f", pc4loadings[1,2])} & \Sexpr{sprintf("%.3f", pc4loadings[1,3])} & \Sexpr{sprintf("%.3f", pc4loadings[1,4])}\\
& Pos. Concept &  \Sexpr{sprintf("%.3f", pc4loadings[2,1])} & \Sexpr{sprintf("%.3f", pc4loadings[2,2])} & \Sexpr{sprintf("%.3f", pc4loadings[2,3])} & \Sexpr{sprintf("%.3f", pc4loadings[2,4])}\\
& Neg. Concept &  \Sexpr{sprintf("%.3f", pc4loadings[3,1])} & \Sexpr{sprintf("%.3f", pc4loadings[3,2])} & \Sexpr{sprintf("%.3f", pc4loadings[3,3])} & \Sexpr{sprintf("%.3f", pc4loadings[3,4])}\\
& Engagement &  \Sexpr{sprintf("%.3f", pc4loadings[4,1])} & \Sexpr{sprintf("%.3f", pc4loadings[4,2])} & \Sexpr{sprintf("%.3f", pc4loadings[4,3])} & \Sexpr{sprintf("%.3f", pc4loadings[4,4])}\\
\hline
\end{tabular}
\caption{Principal Component Analysis with Final Four Item Sets} 
\label{pctable}
\end{table}

>>>>

\begin{figure}[hbtp]
\centering
\includegraphics[keepaspectratio=TRUE,width=.95\textwidth]{ShinyDBSampler/SampleAndSummarizeWithPanelLabels.png}
\caption[\textit{Sample and Summarize} tab screenshot.]{Shiny Database Sampler layout for \textit{Sample and Summarize} tab.} 
\label{fig:samplesummarizetab}
\end{figure}

@@

Once the plotting options are selected in the sidebar panel, the user may click the \textit{Make my Plot!} button to generate and display the plot in the main panel. Since the visualizations are intended only for preliminary exploration, the plots have a default construction and labeling that are not able to be customized. \\

\begin{table}[hbtp]
\centering
\caption{Plot types supported in \textit{Visualize} tab.} 
\begin{tabular}{lll}
\hline
Response Variable & Explanatory Variable & Default Plot \\
\hline
Numerical & None & Histogram \\
Categorical & None & Barchart \\
Numerical & Numerical & Scatterplot \\
Categorical & Categorical & Stacked Barchart \\
Numerical & Categorical & Side-by-side Boxplots \\
\hline
\end{tabular}
\label{tab:defaultplottypes}
\end{table}

\begin{figure}[htbp]
\begin{center}
\includegraphics[keepaspectratio=TRUE,width=.99\textwidth]{ShinyDBSampler/Visualize.png}
\end{center}
\caption[\textit{Visualize} tab screenshot.]{Shiny Database Sampler layout for \textit{Visualize} tab.} 
\label{fig:visualizetab}
\end{figure}

@@

After completing the lab assignment, students were asked to respond to a survey consisting of twelve statements (referred to as \textit{items} in the following, see Table~\ref{tab:surveyquestions} for an overview). For each statement, students were asked for feedback on their level of agreement on a Likert scale from strongly disagree to strongly agree. 
The twelve items were designed to assess student opinion within three topics of four items each: ease of use, connection to sampling concepts, and engagement with the census data.  We will refer to these as the \textit{Ease}, \textit{Concept} and \textit{Engagement} item sets.  For each group of four items, two were worded positively and two were worded negatively. Introducing negation with half of the items was done to reduce the response bias associated with \textit{acquiescence} -- the tendency to respond positively irrespective of the item content due \citep{Furnham1986}. Responses were scored as -2 (strongly disagree), -1 (disagree), 0 (neutral), 1 (agree), 2 (strongly agree).  Responses for negatively worded items were reverse-scored for the purposes of analysis.\\

\begin{table}[hbtp]
\centering
\caption[Survey items and response summaries.]{Survey items and response summaries \textit{after} reverse-scoring (RS).}
\small
% \scalebox{0.9}{
\begin{tabular}{ccp{3.4in}crr}
  \hline
Topic Set & ID & Item & Polarity & Mean & SD\\[3pt]
  \hline
Ease & 1 & \it I found the web tool easy to use & $+$ & \Sexpr{sprintf("%.2f", sumstats[1,2])} & \Sexpr{sprintf("%.2f", sumstats[1,3])}\\  
 & 2 & \it The layout of the web tool was intuitive & $+$  & \Sexpr{sprintf("%.2f", sumstats[4,2])} & \Sexpr{sprintf("%.2f", sumstats[4,3])}\\  
 & 3 & \it Using the web tool was difficult & $-$  & \Sexpr{sprintf("%.2f", sumstats[7,2])}& \Sexpr{sprintf("%.2f", sumstats[7,3])}\\  
 & 4 & \it Learning to use the web tool was hard & $-$  & \Sexpr{sprintf("%.2f", sumstats[10,2])} & \Sexpr{sprintf("%.2f", sumstats[10,3])}\\  [3pt]
  \hline
Concept & 1 & \it The web tool helped me understand sampling concepts & $+$  & \Sexpr{sprintf("%.2f", sumstats[2,2])}& \Sexpr{sprintf("%.2f", sumstats[2,3])}\\   
& 2 & \it I understand sampling ideas less after using the web tool & $-$  & \Sexpr{sprintf("%.2f", sumstats[5,2])}& \Sexpr{sprintf("%.2f", sumstats[5,3])}\\  
& 3 & \it Sampling techniques are clearer after using the web tool & $+$  & \Sexpr{sprintf("%.2f", sumstats[8,2])}& \Sexpr{sprintf("%.2f", sumstats[8,3])}\\  
& 4 & \it The web tool made me less sure how to randomly sample & $-$  & \Sexpr{sprintf("%.2f", sumstats[11,2])}& \Sexpr{sprintf("%.2f", sumstats[11,3])}\\  
   \hline
Engagement & 1 & \it I did not enjoy working with the Census data & $-$  & \Sexpr{sprintf("%.2f", sumstats[3,2])} & \Sexpr{sprintf("%.2f", sumstats[3,3])}\\  
& 2 & \it I thought the Census data was boring & $-$   & \Sexpr{sprintf("%.2f", sumstats[6,2])}& \Sexpr{sprintf("%.2f", sumstats[6,3])}\\  
& 3 & \it Knowing that the Census data was from real people made it more interesting & $+$  & \Sexpr{sprintf("%.2f", sumstats[9,2])}& \Sexpr{sprintf("%.2f", sumstats[9,3])}\\   
& 4 & \it I liked analyzing the Census data & $+$  & \Sexpr{sprintf("%.2f", sumstats[12,2])} & \Sexpr{sprintf("%.2f", sumstats[12,3])}\\   
  \hline
\end{tabular}
% }
\label{tab:surveyquestions}
\end{table}

@@

<<fluctuationDiagrams,echo=F, message=FALSE, include=T,eval=T,fig.width=12, fig.height=6, out.width='.98\\linewidth', fig.pos='hbt',fig.align='center',tidy=F, cache=TRUE, fig.show='hold', fig.cap='Fluctuation diagrams of all item pairs within topic sets.', fig.scap='Fluctuation diagrams of item pairs within topic sets.'>>=

@@

\begin{table}[hbtp]
\centering
\caption{Extended scale for Cronbach's $\alpha$  (George and Mallery, 2003).} 
\begin{tabular}{ll}
\hline 
Internal Consistency & Range \\
\hline
Excellent &  $[ 0.9 , 1.0 ]$ \\
Good & $[ 0.8 , 0.9 )$ \\
Acceptable & $[ 0.7 , 0.8 ) $\\
Questionable & $[ 0.6 , 0.7 )$ \\
Poor & $[ 0.5 , 0.6 )$ \\
Unacceptable & $( -\infty, 0.5 )$ \\
\hline
\end{tabular}
\label{GMAlphaScale}
\end{table}

@@

Table~\ref{cronbachstuff} displays the point estimates and 95\% central bootstrap intervals for Cronbach's $\alpha$ for each item set from the student survey.  The intervals were created using quantiles of the Cronbach's $\alpha$ values from 10,000 bootstrap resamples. 
 The results indicate modest levels of internal consistency for Ease and Engagement item sets, and a lower level for the Concept item set. This is in agreement with the findings based on the fluctuation diagrams in Figure~\ref{fig:fluctuationDiagrams}. \\

@@

% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Tue Nov 04 15:05:46 2014
\begin{table}[hbtp]
\centering
\caption[Cronbach's $\alpha$ estimates for item sets.]{Cronbach's $\alpha$ estimates for each item set with 95\% central bootstrap confidence interval based on 10,000 bootstrap samples.} 
\begin{tabular}{lrc}
  \hline
Set & Estimate & 95\% Confidence Interval \\ 
  \hline
Ease & 0.70 & (0.613 , 0.759)\\ 
  Concept & 0.53 & (0.410 , 0.637) \\ 
  Engagement & 0.72 & (0.643 , 0.776) \\ 
   \hline
\end{tabular}
\label{cronbachstuff}
\end{table}

@@

<<OverallSetBars,echo=F, message=FALSE, include=T ,eval=T,fig.width=9, fig.height=2.5, out.width='.99\\linewidth', fig.pos='hbt',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap= 'Item set response distributions by polarity.'>>=
# meltdatNoNA$value <- as.numeric(as.character(meltdatNoNA$value))
# qu.summary <- ddply(meltdatNoNA, .(set, question.in.set, value), summarize,
# n=length(value) )
# qu.summary$polarity <- rep(c("Pos","Pos","Neg","Neg","Pos","Neg","Pos","Neg","Neg","Neg","Pos","Pos"),each=5)
# qu.summary$glyphShift <- .05
# qu.summary$glyphShift[qu.summary$polarity=="Neg"] <- -.05
# 
# qu.means <- ddply(qu.summary, .(set, value, polarity), summarize, means=mean(n))
# 
# #write.csv(qu.means, "quMeans.csv",row.names=FALSE)
qu.means <- read.csv("quMeans.csv",header=T)

@@

\begin{table}[hbtp]
\centering
\caption[Principal component analysis on topic/polarity item pairs.]{Summary statistics from principal component analysis with six topic/polarity item pairs.} 
\begin{tabular}{lrrrrrrr}
\hline 
& Principal Component & 1 & 2 & 3 & 4 & 5 & 6 \\ 
\hline
Variances \hspace{.1in} & Prop. of Var & \Sexpr{sprintf("%.3f", pc6vars[1,2])} & \Sexpr{sprintf("%.3f", pc6vars[2,2])}  & \Sexpr{sprintf("%.3f", pc6vars[3,2])} & \Sexpr{sprintf("%.3f", pc6vars[4,2])} & \Sexpr{sprintf("%.3f", pc6vars[5,2])} & \Sexpr{sprintf("%.3f", pc6vars[6,2])}  \\
& Cumu. Prop. of Var & \Sexpr{sprintf("%.3f", pc6vars[1,3])} & \Sexpr{sprintf("%.3f", pc6vars[2,3])}  & \Sexpr{sprintf("%.3f", pc6vars[3,3])} & \Sexpr{sprintf("%.3f", pc6vars[4,3])} & \Sexpr{sprintf("%.3f", pc6vars[5,3])} & \Sexpr{sprintf("%.3f", pc6vars[6,3])}  \\
\hline
Loadings & Pos. Ease & \Sexpr{sprintf("%.3f", pc6loadings[1,1])} & \Sexpr{sprintf("%.3f", pc6loadings[1,2])} & \Sexpr{sprintf("%.3f", pc6loadings[1,3])} & \Sexpr{sprintf("%.3f", pc6loadings[1,4])} & \Sexpr{sprintf("%.3f", pc6loadings[1,5])} & \Sexpr{sprintf("%.3f", pc6loadings[1,6])}\\
& Neg. Ease &  \Sexpr{sprintf("%.3f", pc6loadings[2,1])} & \Sexpr{sprintf("%.3f", pc6loadings[2,2])} & \Sexpr{sprintf("%.3f", pc6loadings[2,3])} & \Sexpr{sprintf("%.3f", pc6loadings[2,4])} & \Sexpr{sprintf("%.3f", pc6loadings[2,5])} & \Sexpr{sprintf("%.3f", pc6loadings[2,6])}\\
& Pos. Concept &  \Sexpr{sprintf("%.3f", pc6loadings[3,1])} & \Sexpr{sprintf("%.3f", pc6loadings[3,2])} & \Sexpr{sprintf("%.3f", pc6loadings[3,3])} & \Sexpr{sprintf("%.3f", pc6loadings[3,4])} & \Sexpr{sprintf("%.3f", pc6loadings[3,5])} & \Sexpr{sprintf("%.3f", pc6loadings[3,6])}\\
& Neg. Concept &  \Sexpr{sprintf("%.3f", pc6loadings[4,1])} & \Sexpr{sprintf("%.3f", pc6loadings[4,2])} & \Sexpr{sprintf("%.3f", pc6loadings[4,3])} & \Sexpr{sprintf("%.3f", pc6loadings[4,4])} & \Sexpr{sprintf("%.3f", pc6loadings[4,5])} & \Sexpr{sprintf("%.3f", pc6loadings[4,6])}\\
& Pos. Engaged &  \Sexpr{sprintf("%.3f", pc6loadings[5,1])} & \Sexpr{sprintf("%.3f", pc6loadings[5,2])} & \Sexpr{sprintf("%.3f", pc6loadings[5,3])} & \Sexpr{sprintf("%.3f", pc6loadings[5,4])} & \Sexpr{sprintf("%.3f", pc6loadings[5,5])} & \Sexpr{sprintf("%.3f", pc6loadings[5,6])}\\
& Neg. Engaged &  \Sexpr{sprintf("%.3f", pc6loadings[6,1])} & \Sexpr{sprintf("%.3f", pc6loadings[6,2])} & \Sexpr{sprintf("%.3f", pc6loadings[6,3])} & \Sexpr{sprintf("%.3f", pc6loadings[6,4])} & \Sexpr{sprintf("%.3f", pc6loadings[6,5])} & \Sexpr{sprintf("%.3f", pc6loadings[6,6])}\\
\hline
\end{tabular}
\label{pc6table}
\end{table}

<<PCA6plot,echo=F, message=FALSE, include=T,eval=T,fig.width=5, fig.height=5, out.width='.5\\linewidth', fig.pos='hbt',fig.align='center',tidy=F, cache=TRUE, fig.show='hold', fig.cap='Item pair loadings on Components 2 and 3 from the principal component analysis on six topic/polarity item pairs.', fig.scap='Principal component analysis loadings on topic/polarity item pairs.'>>=
load23dat <- data.frame(pc6loadings[,2:3], compname = row.names(pc6loadings))
require(grid)
ggplot(data=load23dat)+ 
  geom_hline(x=0, colour="darkgray") + geom_vline(y=0, colour="darkgray")+
  geom_point(aes(x=Comp.2, y=Comp.3),size=3.5) + xlim(c(-.73,.6)) + 
  theme_bw() + xlab("Component 2") + ylab("Component 3") +
  geom_segment(aes(x = 0, y = 0, xend = Comp.2, yend = Comp.3), colour="grey40") + 
  geom_text(aes(x=0.825*Comp.2, y=0.825*Comp.3, label=compname)) +coord_fixed() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="cm"))
@

@@

The next major consideration is whether the item sets are truly measuring different latent topics, and therefore can be view as non-redundant. The ability of the survey to separately measure the topics of Ease, Concepts and Engagement can be assessed through the orthogonality of the responses from different item sets. To check the orthogonality of the sets we conduct another principal component analysis; this time on the average responses for each student from the four item sets -- Ease, Positive Concept, Negative Concept and Engagement.  Items sets will be considered highly orthogonal if the principle component analysis cannot reduce the dimensionality from the four sets. \\

@@

\begin{table}[hbtp]
\centering
\caption[Principal component analysis with final four item sets.]{Principal component analysis with final four item sets.} 
\begin{tabular}{lrrrrr}
\hline 
& Principal Component & 1 & 2 & 3 & 4 \\ 
\hline
Variances \hspace{.1in} & Prop. of Var & \Sexpr{sprintf("%.3f", pc4vars[1,2])} & \Sexpr{sprintf("%.3f", pc4vars[2,2])}  & \Sexpr{sprintf("%.3f", pc4vars[3,2])} & \Sexpr{sprintf("%.3f", pc4vars[4,2])}  \\
& Cumu. Prop. of Var & \Sexpr{sprintf("%.3f", pc4vars[1,3])} & \Sexpr{sprintf("%.3f", pc4vars[2,3])}  & \Sexpr{sprintf("%.3f", pc4vars[3,3])} & \Sexpr{sprintf("%.3f", pc4vars[4,3])}  \\
\hline
Loadings & Ease & \Sexpr{sprintf("%.3f", pc4loadings[1,1])} & \Sexpr{sprintf("%.3f", pc4loadings[1,2])} & \Sexpr{sprintf("%.3f", pc4loadings[1,3])} & \Sexpr{sprintf("%.3f", pc4loadings[1,4])}\\
& Pos. Concept &  \Sexpr{sprintf("%.3f", pc4loadings[2,1])} & \Sexpr{sprintf("%.3f", pc4loadings[2,2])} & \Sexpr{sprintf("%.3f", pc4loadings[2,3])} & \Sexpr{sprintf("%.3f", pc4loadings[2,4])}\\
& Neg. Concept &  \Sexpr{sprintf("%.3f", pc4loadings[3,1])} & \Sexpr{sprintf("%.3f", pc4loadings[3,2])} & \Sexpr{sprintf("%.3f", pc4loadings[3,3])} & \Sexpr{sprintf("%.3f", pc4loadings[3,4])}\\
& Engagement &  \Sexpr{sprintf("%.3f", pc4loadings[4,1])} & \Sexpr{sprintf("%.3f", pc4loadings[4,2])} & \Sexpr{sprintf("%.3f", pc4loadings[4,3])} & \Sexpr{sprintf("%.3f", pc4loadings[4,4])}\\
\hline
\end{tabular}
\label{pctable}
\end{table}
